Composable Generalized Consensus

### 2015-05-14 paper/paper.tex
Work on the specification of CGC. Noticed that there is a problem: if learned histories are required to form a chain then the algorithms optimizing for the execution of commuting commands do not implement the spec. Fixed it in the paper but there clearly is no time to fix the TLA+ specs and even less the Isabelle theories.

Command histories can be built from the empty history using the $\star$
operator, which appends a sequence of commands $cs$ to a history $h$, obtaining
the history $h \star cs$. We can then define define an ordering relation among
histories: $h_1 \leq h_2$ when there exists a sequence $cs$ such that $h_2 =
h_1 \star cs$. 
The maximum over a set of
histories $H$ is denoted by $Max\left( H \right)$ and $Max\left( \aset{}
\right)$ is the empty history.
Histories have the property that every finite set $H$ of 
histories has a \emph{greatest lower bound}, noted $GLB\left(H\right)$.

@@

To specify Chain Agreement, we need the notion of ordering of sequences and of
maximum sequence.  The prefix order on sequences is denoted by the operator
$\leq$ ($s_1 \leq s_2$ means that $s_1$ is a prefix of $s_2$). 
%Using the operator $\leq$, we define $\geq$, $<$, and $>$ as expected. 
Note that $\leq$ is a partial order on sequences. The maximum over a set of
sequences $S$ is denoted by $Max\left( S \right)$ and $Max\left( \aset{}
\right)$ is the empty sequence.

We specify Chain Agreement using the state variables $crashed$, $proposed$, containing the proposed commands,
and, for every learner $l$,  $learned\left[ l \right]$, a set of
sequences of commands. Initially, $crashed$, $proposed$ and, for every learner $l$,
$learned\left[ l \right]$ are empty sets. 
Remember that any transition involving a server $x$ is always disabled when $x\in crashed$.

The next-state relation consists of three types of transitions:
$Crash\left( x \right)$ transitions; $Propose\left( l, c
\right)$ transitions, for a learner $l$ and a command $c$, in which
$c$ is inserted in the set $proposed$; $Learn\left( l,cs \right)$
transitions, for a learner $l$ and a sequence of commands $cs$, in
which $cs$ is inserted in $learned\left[ l \right]$. 
A $Learn\left( l, cs \right)$ transition is guarded by the following
conditions:
\begin{compactitem}
    \item[\textbf{Agreement}:] The elements of the set $\aset{learned'\left[ l
      \right] : l \in Learners}$, the learned sequences after the transition,
      form a chain in the partial order $\leq$.
    \item[\textbf{Validity}:] Every learned sequence contains only
        proposed commands.  
    \item[\textbf{Irrevocability}:] The new learned sequence
        $cs$ is a strict extension of $Max\left(learned\left[ l
        \right]\right)$. 
        %\\ $cs > Max\left(learned\left[ l \right]\right)$.
\end{compactitem}
Finally, we add, for every learner $l$, one weak fairness condition
\cite{Lamport02SpecifyingSystems} on the transitions $Learn\left( l,
cs \right)$. The weak fairness conditions ensure that every correct
learner eventually learns about every proposed command.

Thanks to the properties of Chain Agreement, the clients, observing
their input commands and the received responses, see a
\emph{linearizable} execution
\cite{HerlihyWing90LinearizabilityCorrectnessConditionConcurrentObjects}.
Moreover, the service remains available as long as at least one
proposer and one learner are correct.

@@

\subsection{Histories}
\label{sec:histories}

ACA can be implemented more efficiently by substituting \emph{command histories} for sequences of commands, as in Generalized Consensus \cite{Lamport05GeneralizeConsensus}.

We say that two commands $c_1$ and $c_2$ \emph{commute} when, in every sequential
execution of the service, reversing the order of $c_1$ and $c_2$ does
not change any of the outputs, including those given to $c_1$ and
$c_2$.

A command history, abbreviated history, is a data structure that
represents a set of sequences which are the same modulo the reordering
of commuting commands, and in which duplicate commands do not appear.
All the sequences represented by a given history contain the same
commands and determine the same outputs for each command.  Therefore,
to implement a replicated service, it is sufficient that learners
learn command histories instead of sequences.  For a precise
definition of command histories, we refer the reader to
\cite{Lamport05GeneralizeConsensus}. Our Isabelle/HOL formalization
(\cref{sec:isaproofs}) abstracts
over the concrete representation of histories, assuming only their necessary
properties. Command histories were first
introduced in trace theory \cite{Mazurkiewicz84Semantics}.

A command history can be built by appending a sequence of commands
$cs$ to a history $h$, obtain the history $h \star cs$.  The major
advantage that histories have over sequences is that if a sequence
$cs_1$ is obtained by reordering the commuting commands of $cs_2$,
then $h \star cs_1 = h \star cs_2$. Therefore, in  implementations of
ACA, servers can agree on a history even though
they chose different orders for some commuting commands.

Command histories share the four properties of sequences that are
necessary and sufficient to substitute histories for sequences in the
definition of Chain Agreement and ACA.
\begin{compactitem}
    \item[\textbf{Distinctness}:] No duplicate command appears in a history. 
    \item[\textbf{Ordering}:] Command histories form a partial order:
        $h_1 \leq h_2$ when there exists a sequence $s$ such that $h_2
        = h_1 \star s$.
    \item[\textbf{GLB}:] Every finite set of command histories has a
        \emph{greatest lower bound}.
%    \item[\textbf{LUB}:] Every finite set of compatible command
%        histories has a \emph{least upper bound}, where a set of
%        command histories is compatible when they have a common upper
%        bound.
    \item[\textbf{Consistency}:] If $h_1 \leq h_2 \leq h_3$ and $h_3 =
        h_1 \star s$ then there exists $s'$ and $s''$ where every
        request of $s'$ and $s''$ appears in $s$ and such that $h_2 =
        h_1 \star s'$ and $h_3 = h_2 \star s''$.
\end{compactitem}
The optimized version of ACA is obtained by
simply substituting histories for sequences in the definition of ACA.
The Ordering, GLB, and Consistency properties of histories ensure that
\cref{thm:comp} still holds: the Isabelle/HOL proof of \cref{thm:comp} (see
\cref{sec:isaproofs}) relies only on the four properties above.

>>>>

Command histories can be built from the empty history using the $\star$
operator, which appends a sequence of commands $cs$ to a history $h$, obtaining
the history $h \star cs$. We can then define define an ordering relation among
histories: $h_1 \leq h_2$ when there exists a sequence $cs$ such that $h_2 =
h_1 \star cs$. 
The maximum over a set of
histories $H$ is denoted by $Max\left( H \right)$ and $Max\left( \aset{}
\right)$ is the empty history.
Histories have the property that every finite set $H$ of 
histories has a \emph{greatest lower bound}, noted $GLB\left(H\right)$.
Finally, we say that a set of histories $H$ is compatible when all the histories in $H$ have a common upper bound $h$ such
that $h$ can be built using only commands found in the histories in $H$.

@@

We specify Generalized Consensus using the state variables $crashed$, $proposed$, containing the proposed commands,
and, for every learner $l$,  $learned\left[ l \right]$, a set of
sequences of commands. Initially, $crashed$, $proposed$ and, for every learner $l$,
$learned\left[ l \right]$ are empty sets. 
Remember that any transition involving a server $x$ is always disabled when $x\in crashed$.

The next-state relation consists of three types of transitions:
$Crash\left( x \right)$ transitions; $Propose\left( l, c
\right)$ transitions, for a learner $l$ and a command $c$, in which
$c$ is inserted in the set $proposed$; $Learn\left( l,cs \right)$
transitions, for a learner $l$ and a sequence of commands $cs$, in
which $cs$ is inserted in $learned\left[ l \right]$. 
A $Learn\left( l, cs \right)$ transition is guarded by the following
conditions:
\begin{compactitem}
    \item[\textbf{Agreement}:] The set of learned histories after the transition, $\aset{learned'\left[ l
      \right] : l \in Learners}$, is compatible.
    \item[\textbf{Validity}:] Every learned sequence contains only
        proposed commands.  
    \item[\textbf{Irrevocability}:] The new learned sequence
        $cs$ is a strict extension of $Max\left(learned\left[ l
        \right]\right)$. 
        %\\ $cs > Max\left(learned\left[ l \right]\right)$.
\end{compactitem}

Thanks to the properties of Generalized Consensus, the clients, observing
their input commands and the received responses, see a
\emph{linearizable} execution
\cite{HerlihyWing90LinearizabilityCorrectnessConditionConcurrentObjects}.
Moreover, the service remains available as long as at least one
proposer and one learner are correct.

@@
